---
title: "Many Local Maxima and Modeling Beetle Data"
subtitle: "HW 4 of STAT 5361 Statistical Computing"
author: Biju Wang^[<bijuwang@uconn.edu>]
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
header-includes: 
  \usepackage{float}
  \floatplacement{figure}{H}
output: 
  pdf_document:
    number_sections: true
---

# Many Local Maxima
## 

The likelihood function is
$$L(\theta)=\prod^{n}_{i=1}\frac{1-\cos(X_{i}-\theta)}{2\pi}$$
Thus, the log likelihood fucntion is
$$l(\theta)=\sum^{n}_{i=1}\log[1-\cos(X_{i}-\theta)]-n\log2\pi$$
The following plot is the curve of log likelihood function
```{r, echo = TRUE, fig.asp=.32, fig.pos = "H",fig.cap = "Log Likelihood Function vs. $\\theta$"}
sample <- c(3.91, 4.85, 2.28, 4.06, 3.70, 4.04, 5.46, 3.53, 2.28, 1.96,
       2.53, 3.88, 2.22, 3.47, 4.82, 2.46, 2.99, 2.54, 0.52)
log_sum <- function(x, sample){
  log_sum <- 0
  for (i in 1:length(sample)) {
    log_sum <- log_sum + log(1 - cos(sample[i] - x)) - log(2 * pi)
  }
  log_sum
}
library("ggplot2")
ggplot(data.frame(x = c(-pi, pi)), aes(x = x)) + 
stat_function(fun = function(x) log_sum(x, sample)) + 
labs(x = expression("Values of"~theta), y = expression("Log Likelihood Function"~l(theta))) + 
theme(plot.title = element_text(hjust = 0.5)) + 
ggtitle(expression("Log Likelihood Function vs."~theta))
```

##

\begin{align*}
E(X|\theta) & =\int^{2\pi}_{0}x\frac{1-\cos(x-\theta)}{2\pi}dx\\
            & =\frac{1}{2\pi}\left[\int^{2\pi}_{0}xdx-\int^{2\pi}_{0}x\cos(x-\theta)dx\right]\\
            & =\frac{1}{2\pi}\left(2\pi^{2}+2\pi\sin\theta\right)\\
            & =\pi+\sin\theta
\end{align*}
Let $E(X|\theta)$ equal sample mean $\bar{X}_{n}$
$$\pi+\sin\theta=\bar{X}_{n}$$
we could obtain the method-of-moments estimator of $\theta$ is $\tilde{\theta}_{n}=0.095$ or $3.046$.

##

The first derivative is
$$l'(\theta)=\sum^{n}_{i=1}\frac{-\sin(X_{i}-\theta)}{1-\cos(X_{i}-\theta)}$$

The second derivative is
$$l''(\theta)=\sum^{n}_{i=1}\frac{1}{\cos(X_{i}-\theta)-1}$$
```{r, message = FALSE}
sample <- c(3.91, 4.85, 2.28, 4.06, 3.70, 4.04, 5.46, 3.53, 2.28, 1.96,
       2.53, 3.88, 2.22, 3.47, 4.82, 2.46, 2.99, 2.54, 0.52)

dev1_log_sum <- function(x){
  dev1_log_sum <- 0
  for (i in 1:length(sample)) {
    dev1_log_sum <- dev1_log_sum - sin(sample[i] - x)/(1 - cos(sample[i] - x))
  }
  dev1_log_sum
}

dev2_log_sum <- function(x){
  dev2_log_sum <- 0
  for (i in 1:length(sample)) {
    dev2_log_sum <- dev2_log_sum + 1/(cos(sample[i] - x) - 1)
  }
  dev2_log_sum
}

newton.raphson <- function(init, fun, fun.dev, maxiter = 100, tol = .Machine$double.eps^0.2){
  x <- init
  for (i in 1:maxiter) {
    x1 <- x -fun(x)/fun.dev(x)
    if(abs(x1 - x) < tol) break
    x <- x1
  }
  if(i == maxiter)
    message("Reached the maximum iteration!")
  
  return(data.frame(root = x1, iter = i))
}

init <- c(asin(mean(sample) - pi), pi - asin(mean(sample) - pi))
res <- data.frame(init = init, root = rep(NA, length(init)))
for (i in 1:length(init)) {
  res$root[i] <- newton.raphson(init[i], dev1_log_sum, dev2_log_sum)$root
}

res_trans <- t(as.matrix(res))
rownames(res_trans) <- c("Initial Values", "Roots")

knitr::kable(res_trans, booktabs = TRUE,
             caption = "Initial Values and Roots")

```
##

The outcomes are in the following table

```{r, message = FALSE}
init <- c(-2.7, 2.7)
res <- data.frame(init = init, root = rep(NA, length(init)))
for (i in 1:length(init)) {
  res$root[i] <- newton.raphson(init[i], dev1_log_sum, dev2_log_sum)$root
}

res_trans <- t(as.matrix(res))
rownames(res_trans) <- c("Initial Values", "Roots")

knitr::kable(res_trans, booktabs = TRUE,
             caption = "Initial Values and Roots")
```
##

```{r, fig.asp=.5, fig.pos = "H", fig.cap = "Scatter Plot of Roots vs. Initial Values"}
options(digits = 6)
init <- seq(-pi, pi, length = 200)
res <- data.frame(init = init, root = rep(NA, length(init)))
for (i in 1:length(init)) {
  res$root[i] <- newton.raphson(init[i], dev1_log_sum, dev2_log_sum)$root
}

res <- round(res, 5)
res_trans <- t(as.matrix(res))
rownames(res_trans) <- c("Initial Values", "Roots")

library("pander")
library("ggplot2")
pander(res_trans, split.table = 120, style = 'rmarkdown', caption = "Initial Values and Roots")

ggplot(res, aes(x = init, y = root)) + geom_point() + 
scale_x_continuous(breaks = round(seq(min(res$init), max(res$init), by = 1),1)) +
labs(x = "Initial Values", y = "Roots from Newton-Raphson") + 
theme(plot.title = element_text(hjust = 0.5)) + 
ggtitle("Scatter Plot of Roots vs. Initial Values")

table <- as.data.frame(table(res[,2]))
table <- t(as.matrix(table))
table <- rbind(table, cumsum(table[2,]))
rownames(table) <- c("Roots", "Amount", "Cumulative Amount")
pander(table, split.table = 120, style = 'rmarkdown', caption = "Roots, Amount and Cumulative Amount")

```

From the table above, we can partition the $200$ initial points into $21$ separate groups with each group corresponding to a separate unique outcome of the optimization. According to order number of initial values, the $21$ groups are
\begin{table}[H]
\begin{center}
\begin{tabular}{cccccc}
$[1, 11]$ & $[12, 13]$ & $[14, 18]$ & $[19, 24]$ & $[25, 25]$ & $[26, 28]$\\
$[29, 29]$ & $[30, 30]$ & $[31, 54]$ & $[55, 55]$ & $[56, 69]$ & $[70, 74]$\\
$[75, 116]$ & $[117, 162]$ & $[163, 170]$ & $[171, 171]$ & $[172, 172]$ & $[173, 178]$\\
$[179, 180]$ & $[181, 195]$ & $[196, 200]$
\end{tabular}
\end{center}
\caption{Groups of Initial Values for the Same Root}
\end{table}

# Modeling Beetle Data
## Least Square Method
The solution for the differential equation is not unique unless we give a initial condition which will determine $N_{0}$. In the following solutions, we assume $N_{0}=2$. This is reasonable since at time $0$ the population size is $2$. We need to minimize
$$g(r, K)=\sum^{n}_{i=1}\left[N_{i}-\frac{2K}{2+(K-2)e^{-rt_{i}}}\right]^{2}$$
We use Gauss-Newton approach to address this optimization problem, this is also the default method of the function nlxb in package nlmrt. The initial values for both $K$ and $r$ are $1$. The roots and contor plot are below 

```{r, fig.asp=.5, fig.pos = "H", fig.cap = "Contour Plot for the Objective Function"}
beetles <- data.frame(
    days    = c(0,  8,  28,  41,  63,  69,   97, 117,  135,  154),
    beetles = c(2, 47, 192, 256, 768, 896, 1120, 896, 1184, 1024))

goal <- function(r, K){
  goal <- 0
  for (i in 1:nrow(beetles)) {
    goal <- goal + (beetles[i,2]-(2 * K)/(2 + (K - 2) * exp(-r * beetles[i,1])))^2
  }
  goal
}

library("ggplot2")
r <- seq(1, 2000, 0.1)
K <- seq(1, 2000, 0.1)
data <- data.frame(r = r, K = K)
data$z <- with(data, goal(r, K))
ggplot(data, aes(r, K, z))+ geom_point(aes(colour = z)) + stat_density2d() +
theme(plot.title = element_text(hjust = 0.5)) + 
ggtitle("Contour Plot for the Objective Function")

library("nlmrt")
mod.data <- data.frame(x = beetles[,1], y = beetles[,2])
nlmod <- nlxb(y ~ 2 * K/(2 + (K - 2) * exp(-r * x)), data = mod.data, start = c(K = 1, r = 1))
nlmod$coefficients
```
## MLE Method
Since we assume $\log N_{t}\sim N(\log f(t), \sigma^{2})$, the likelihood function is
$$L(r, K, \sigma^{2})=\prod^{n}_{i=1}\frac{1}{\sqrt{2\pi\sigma^{2}}}\exp(-\frac{(\log N_{i}-\log\frac{2K}{2+(K-2)\exp(-rt_{i})})^{2}}{2\sigma^{2}})$$
Thus, the log likelihood function is
$$l(r, K, \sigma^{2})=-\frac{n}{2}\log(2\pi\sigma^{2})-\sum^{n}_{i=1}\frac{(\log N_{i}-\log\frac{2K}{2+(K-2)\exp(-rt_{i})})^{2}}{2\sigma^{2}}$$
We would like to maximize the log likelihood function regarding to $r, K, \sigma^{2}$. Since the objective function is very complex, directly using Newton-Raphson method is not appropriate. We instead use Quasi-Newton method, specifically, we use BFGS method in package optimx or optim. The point estimations and variance for it are shown below

```{r}
library("optimx")
goal <- function(x){
  goal <- 0
  r <- x[1]
  K <- x[2]
  sigma2 <- x[3]
  for (i in 1:nrow(beetles)) {
    goal <- goal - log(2 * pi * sigma2)/2 - 
      (log(beetles[i,2]) - log((2 * K)/(2 + (K - 2) * exp(-r * beetles[i,1]))))^2/(2 * sigma2)
  }
  goal
}

BFGSmod <- optimx(c(800, 800, 70), fn = goal, gr = NULL, method = "BFGS", hessian = TRUE)
coef.BFGSmod <- coef(BFGSmod)
colnames(coef.BFGSmod) <- c("r", "K", "sigma^2")
BFGSmod
coef.BFGSmod
BFGSmod1 <- optim(c(800, 800, 70), fn = goal, gr = NULL, method = "BFGS", hessian = TRUE)
BFGSmod1$hessian
MASS::ginv(-BFGSmod1$hessian)
```

